Using reference models in variable selection \vspace{.1in}

### 2022-01-19 paper/ref_approach.tex
more edits

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper, we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability, leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model, leading to projection predictive variable selection approach. Alternatively, reference models may also be used in an ad-hoc manner in combination with common variable selection methods.  In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection. Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods, independently of whether or not they use reference models.
\end{abstract}

@@

In data-based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward
regression \citep{venables2013modern,harrell2015regression}, the
observed empirical data distribution is utilised as a proxy of future
data, usually in combination with cross-validation or information
criteria to provide estimates of out-of-sample predictive
performance.  In contrast, reference model based methods approximate
the future data generation mechanism using the predictive distribution
of a reference model, which can be, for example, a full-encompassing
model including all variables.
%
We assume all models are wrong, but we assume we have constructed a
model which reflects our beliefs about the future data in the best
possible way and which has passed model checking and criticism
\citep{gelman2020bayesian,gabry2019visualization}.  Using the usual
best practices for constructing the reference model is important, as
using a bad reference model can only lead to selecting a similarly
bad smaller model. If the reference model is considered to have
useful predictions, then the smaller models selected will also have
similar useful predictions.
%
The reference model approach has been used in Bayesian statistics in
some form at least since the seminal work of
\citet{paper:reference_lindley}. For more historical references, see
\citet{vehtari2012survey} and \citet{paper:model_selection}, and for
most recent methodological developments see \citet{paper:projpred}.
%
Examples of useful reference models can be found for example for
small-$n$-large-$p$ regression and logistic regression by
\citet{piironen2017comparison} (with spike-and-slab prior),
\citet{Piironen:2015:Stan_projection} (with horseshoe prior), and
\citet{paper:projpred} (with iterative supervised principal
components), for generalized linear and additive multilevel models by
\citet{Catalina:2020:GAMM_projection}, for regression models with
non-exponential family observation models by
\citet{Catalina+etal:2021:latent_projection}, and for generic
multivariate non-linear regression with higher order interactions by
\citet{Piironen:2016:GPprojection}.

Reference models have been also used in non-Bayesian contexts, in which
\cite{harrell2015regression} describes them as full models that can be
thought of as a ``gold standard'' (for a given application).  For
example, \cite{faraggi2001understanding} deal with the necessity of
identifying interpretable risk groups in the context of survival data
using neural networks, which typically perform very well in terms of
prediction, but whose variables are difficult to be understood in
terms of relevance.  \cite{paul2008preconditioning}, using the term
preconditioning, explore approximating models fitting Lasso or
stepwise regression against consistent estimates $\hat{y}$ of a
reference model instead of the observed responses $y$.

\fede{All these methods can be framed into the family of reference model approaches. 
The common denominator is the use of the reference model posterior predictive information 
instead of simply observed data to guide the selection.}
Whatever the terminology or applied statistical framework, reference
models offer a powerful approach to improving variable selection, as we
will demonstrate in the present paper. \\
\fede{The goal of our work is to study the impact of reference model approaches
disentangling the benefit of the usage of a reference model from the specific selection algorithm.
Specifically, we:}
% The main contributions of our work are:
\fede
{\begin{itemize}
\item disentangle the action of the reference model through a simple and intuitive way to combine any variable selection method with the reference model approach. This allows us to build up a comparison independent of the specific variable selection method;
% We show a simple and intuitive way to combine any variable selection method with the reference model approach.
\item perform extensive numerical experiments to compare variable selection methods with or without using a reference model, both for complete and minimal subset feature selection and assessing the quality of the selection;
\item provide evidence supporting, in particular, the projection predictive approach as a principled way to use reference models in variable selection.
\end{itemize}
}

@@

To motivate the further discussion and experiments, we start by a simple
variable selection example using body fat data by \citet{johnson1996fitting}.
We compare the projective prediction approach (\emph{projpred},
\citet{paper:projpred}) which uses a reference model, and classic
stepwise backward regression (\emph{steplm}).
The experiments are implemented in \texttt{R} \citep{Rcore2018}.

@@

Both projpred and steplm compare a large number of models using either
forward or backward search, which can lead to selection induced
overfitting, but even with 100 variables, projpred is able to select a
submodel with similar performance as the full model. In this example, the two
 compared methods also differ in other aspects than the usage of a reference 
 model, such as that projpred
uses Bayesian inference and steplm uses maximum likelihood estimation. 
However, as we show in the present paper, one of the primary difference 
with respect to quality of the variable selection is indeed whether or 
not a reference model is applied, and the performance of steplm can
also be improved by using a reference model.

@@

\fede{In general, there is not restriction on the type of model the reference models 
should be and a sensible reference model does not need to be Bayesian.
However, Bayesian methods can help in some of those situations
where MLE procedures struggle. 
It is known that as the number of parameters in the model increases, the MLE
estimator is dominated by shrunk estimators 
\citep{stein1956inadmissibility, stein1961estimation, parmigiani2009decision, efron2011tweedie}.
The use of a prior in Bayesian inference automatically incorporates 
some kind of shrinkage in the Bayesian estimator under a given loss function \citep[see, e.g.][]{rockova2012hierarchical}. 
For the sake of our study, we rely on simple data structure which can be well described 
by linear regression models. However, more complex data can arise from applications, for example including hierarchical structures. In these cases, MLE inference tends to become cumbersome, whereas the Bayesian framework provides a natural way to convey uncertainties and make inference through the posterior distribution.}

@@

In Figure \ref{fig:rmse_vs_fdr}, the predictive performance is shown
for different values of $n$ and $\rho$ in terms of RMSE and the false discovery rate (FDR, the ratio of the number of non-relevant selected
variables over the number of selected variables) of the selected submodel,
averaged after 100 data simulations. Use of a reference model reduces RMSE of the selected model for all the methods. projpred has superior FDR and the
smallest RMSE. Using a reference model improves also FDR of steplm significantly. Stepwise
Bayesian linear regression is better than steplm when comparing both methods with
and without a reference model, respectively. However, the minimal subset variable
selection of the Bayesian linear regression is still less reliable than projpred.

@@

Figure \ref{fig:sensitivity_vs_fdr_iterated} shows the average sensitivity on
the vertical axis and the average false discovery rate on the
horizontal axis based on 100 data simulations for the different
combinations of $n$ and $\rho$. The best selection performance is on the
top-left corner of each plot, as it implies the lowest false discovery
rate and the highest sensitivity. We see that regardless of the
applied selection method, the use of a reference model improves the
selection performance, as it reduces the false discovery rate (shifting
to the left) or increases the sensitivity (shifting upwards). In
accordance with what can be expected, the larger data set size ($n$) and the higher the true correlations ($\rho$), the easier the
selection is. Thus, for easier selection scenarios, the benefits of
the reference model are smaller since the raw data already provide
enough information to identify the relevant variables. The iterative
projpred has good false discovery rate in all cases, and the
sensitivity is good except when the number of observations and the
correlation level are small. It performs better than the iterative lasso 
selection in any of the simulated scenarios. Tuning $\alpha$ might lead to improved
selection, yet the main source of poor performance could well be the
miscalibration of the uncertainty of the predictive utility score for
misspecified models with cross-validation \citep{bengio2004no}.

Figure \ref{fig:stability_iterated} shows the estimates of the stability
measure proposed by \cite{paper:stability} with 0.95 confidence
intervals based on 100 simulations. Such a measure takes to account
the variability of the subset of the selected variables at each
simulation (originally at each bootstrap sample), modelling the
selection of each variable as a Bernoulli process. Further details are
available in \cite{paper:stability}. The reference model helps in
improving the stability of the selection: again, the benefits are
bigger when the problem is more difficult (small $n$ and $\rho$). In
addition, we observe less uncertainty in the stability estimates for
the reference approach (i.e., smaller width of the 95\% intervals),
which can be still connected to the overall stability of the
procedure. As in Figure \ref{fig:sensitivity_vs_fdr_iterated}, the iterative
projection does not perform well in the hardest scenarios.

>>>>

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper, we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability, leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model, leading to projection predictive variable selection approach. We analyse how much the great performance of the projection predictive variable is due to the use of reference model and show that other variable selection methods can also be greatly improved by using the reference model as target instead of the original data. In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection.
  %Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods, independently of whether or not they use reference models.
\end{abstract}

@@

In data-based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward
regression \citep{venables2013modern,harrell2015regression}, the
observed empirical data distribution is utilised as a proxy of future
data, usually in combination with cross-validation or information
criteria to provide estimates of out-of-sample predictive
performance.  In contrast, reference model based methods approximate
the future data generation mechanism using the predictive distribution
of a reference model, which can be, for example, a full-encompassing
model including all variables.

We assume all models are wrong, but we assume we have constructed a
model which reflects our beliefs about the future data in the best
possible way and which has passed model checking and criticism
\citep{gelman2020bayesian,gabry2019visualization}.  Using the usual
best practices for constructing the reference model is important, as
using a bad reference model can only lead to selecting a similarly
bad smaller model. If the reference model is considered to have
useful predictions, then the smaller models selected will also have
similar useful predictions.
%
The reference model approach has been used in Bayesian statistics in
some form at least since the seminal work of
\citet{paper:reference_lindley}. For more historical references, see
\citet{vehtari2012survey} and \citet{paper:model_selection}, and for
most recent methodological developments see \citet{paper:projpred}.
%
Examples of useful reference models can be found for example for
small-$n$-large-$p$ regression and logistic regression by
\citet{piironen2017comparison} (with spike-and-slab prior),
\citet{Piironen:2015:Stan_projection} (with horseshoe prior), and
\citet{paper:projpred} (with iterative supervised principal
components), for generalized linear and additive multilevel models by
\citet{Catalina:2020:GAMM_projection}, for regression models with
non-exponential family observation models by
\citet{Catalina+etal:2021:latent_projection}, and for generic
multivariate non-linear regression with higher order interactions by
\citet{Piironen:2016:GPprojection}.

Reference models have been also used in non-Bayesian context.
\cite{harrell2015regression} describes them as full models that can be
thought of as a ``gold standard'' (for a given application).
\cite{faraggi2001understanding} deal with the necessity of
identifying interpretable risk groups in the context of survival data
using neural networks, which typically perform very well in terms of
prediction, but whose variables are difficult to be understood in
terms of relevance.  \cite{paul2008preconditioning}, using the term
preconditioning, explore approximating models fitting Lasso or
stepwise regression against consistent estimates $\hat{y}$ of a
reference model instead of the observed responses $y$.

\fede{All these methods can be framed into the family of reference model approaches. 
The common denominator is the use of the reference model predictive information 
instead of simply observed data to guide the selection.}
Whatever the terminology or applied statistical framework, reference
models offer a powerful approach to improving variable selection, as we
will demonstrate in the present paper.

\fede{The goal of our work is to study the impact of reference model approaches
disentangling the benefit of the usage of a reference model from the specific selection algorithm.
Specifically, we:}
% The main contributions of our work are:
\fede
{\begin{itemize}
\item disentangle the action of the reference model through a simple and intuitive way to combine any variable selection method with the reference model approach. This allows us to build up a comparison independent of the specific variable selection method;
% We show a simple and intuitive way to combine any variable selection method with the reference model approach.
\item perform extensive numerical experiments to compare variable selection methods with or without using a reference model, both for complete and minimal subset feature selection and assessing the quality of the selection;
\item provide evidence supporting, in particular, the projection predictive approach as a principled way to use reference models in variable selection.
\end{itemize}
}

@@

To motivate the further discussion and experiments, we start by a simple
variable selection example using body fat data by \citet{johnson1996fitting}. The same data was used illustrate the variable selection with classic stepwise backward regression by \citet{paper:bodyfat}.
We compare the projective prediction approach (\emph{projpred},
\citet{paper:projpred}) which uses a reference model, and classic
stepwise backward regression (\emph{steplm}).
The experiments are implemented in \texttt{R} \citep{Rcore2018}.

@@

Both projpred and steplm compare a large number of models using either
forward or backward search, which can lead to selection induced
overfitting, but even with 100 variables, projpred is able to select a
submodel with similar performance as the full model. In this example, the two
compared methods also differ in other aspects than the usage of a reference 
model, such as that projpred
uses Bayesian inference and steplm uses maximum likelihood estimation.
%
To separate the effect of using a reference model, we show that
performance of other variable selection methods, including steplm, can
also be improved by using reference models.
% However, as we show in the present paper, one of the primary difference 
% with respect to quality of the variable selection is indeed whether or 
% not a reference model is applied, and the performance of steplm can
% also be improved by using a reference model.

@@

\fede{In general, there is not restriction on the type of model the reference models 
should be and a sensible reference model does not need to be Bayesian.
However, Bayesian methods can help in some of those situations
where MLE procedures struggle. 
It is known that as the number of parameters in the model increases, the MLE
estimator is dominated by shrinkage estimators 
\citep{stein1956inadmissibility, stein1961estimation, parmigiani2009decision, efron2011tweedie}.
The use of a prior in Bayesian inference automatically incorporates 
some kind of shrinkage in the Bayesian estimator under a given loss function \citep[see, e.g.][]{rockova2012hierarchical}. 
For the sake of our study, we rely on simple data structure which can be well described 
by linear regression models. However, more complex data can arise from applications, for example including hierarchical structures. In these cases, MLE inference tends to become cumbersome, whereas the Bayesian framework provides a natural way to convey uncertainties and make inference through the posterior distribution.}

@@

In Figure \ref{fig:rmse_vs_fdr}, the predictive performance is shown
for different values of $n$ and $\rho$ in terms of RMSE and the false discovery rate (FDR, the ratio of the number of non-relevant selected
variables over the number of selected variables) of the selected submodel,
averaged after 100 data simulations. Use of a reference models greatly reduces the overfitting in the selection process, and thus produces much lower test set RMSE of the selected model for all the methods. projpred has the lowest FDR and RMSE. Using a reference model improves also FDR of steplm stepwise Bayesian liena rgegression significantly.

@@

Figure \ref{fig:sensitivity_vs_fdr_iterated} shows the average sensitivity on
the vertical axis and the average false discovery rate on the
horizontal axis based on 100 data simulations for the different
combinations of $n$ and $\rho$. The best selection performance is on the
top-left corner of each plot, as it implies the lowest false discovery
rate and the highest sensitivity. We see that for all tested
selection methods, the use of a reference model improves the
selection performance, as it reduces the false discovery rate (shifting
to the left) and increases the sensitivity (shifting upwards). In
accordance with what can be expected, the larger data set size ($n$) and the higher the true correlations ($\rho$), the easier the
selection is. Thus, for easier selection scenarios, the benefits of
the reference model are smaller since the raw data already provide
enough information to identify the relevant variables. The iterative
projpred has good false discovery rate in all cases, and the
sensitivity is good except when the number of observations and the
correlation level are small. It performs better than the iterative lasso 
selection in all simulated scenarios.
%
For all these variable selection methods, tuning the corresponding
method parameters could affect the balance between FDR and
sensitivity, and thus the methods could made to produce more similar
results. This was not done in this experiment, as the main point was
to show that all methods can perform better with a reference model.
% Tuning $\alpha$ might lead to improved selection, yet the main
% source of poor performance could well be the miscalibration of the
% uncertainty of the predictive utility score for misspecified models
% with cross-validation \citep{Sivula+etal:2020:loo_uncertainty}.

Figure \ref{fig:stability_iterated} shows the estimates of the stability
measure proposed by \cite{paper:stability} with 0.95 confidence
intervals based on 100 simulations. Such a measure takes to account
the variability of the subset of the selected variables at each
simulation (originally at each bootstrap sample), modelling the
selection of each variable as a Bernoulli process. Further details are
available in \cite{paper:stability}. Use of a reference model 
improves the stability of all selection methods. The improvement is 
larger when the problem is more difficult (small $n$ and $\rho$). In
addition, we observe less uncertainty in the stability estimates for
the reference approach (i.e., smaller width of the 95\% intervals),
which can be still connected to the overall stability of the
procedure. As in Figure \ref{fig:sensitivity_vs_fdr_iterated}, the iterative
projection does not perform well in the hardest scenarios.
